---
title: "Biclusters for LEMUR -- a workflow"
author: "Tim"
date: last-modified
format:
  html:
    code-fold: true
    code-summary: "Show code"
    theme: cosmo 
    self-contained: true 
    page-layout: full  
execute:
  warning: false
  message: false
---


```{r}
#| label: setup
#| include: false

# source the protopackage
SOURCE_DIR <- "/Users/herbermann/code/lemur_biclusters/R"

files_to_source <- c(
  "preprocess.R",
  "explore_factors.R",
  "fit_nmf.R",
  "extract_biclusters.R",
  "recruitment.R",
  "test_bicluster.R",
  "consensus_biclustering.R"
)

invisible(lapply(file.path(SOURCE_DIR, files_to_source),source))

library("tidyverse")
library("SingleCellExperiment")
library("lemur")
library("pheatmap")

data("glioblastoma_example_data", package = "lemur")

set.seed(42)
```

# 0. LEMUR standard fit

```{r}
lemur_fit <- lemur(glioblastoma_example_data, design = ~ patient_id + condition, 
             n_embedding = 15, test_fraction = 0.5)
lemur_fit <- align_harmony(lemur_fit)
lemur_fit <- test_de(lemur_fit, contrast = cond(condition = "panobinostat") - cond(condition = "ctrl"))
```

# 1. Preprocess

Some preprocessing steps to crystalise the bicluster structure better.
Options here might be:

- soft/hard max clipping: DE of 2,4 or 8 logfolds are similarly relevant and can be mapped close to each other

- LEMUR fit induces dense noise around 0. It may be beneficial to clip that noise away to increase performance and avoid biclustering to noise from the original fit.

Here, we take the DE predicted from LEMUR, $\Delta Y$, and apply $\Delta \hat Y = \frac{1}{w} \mathrm{tanh}(w \Delta Y)$ where $w = 1/Q_{95\%}$. (The used function already has placerholder for different transformations, but those are not yet implemented). The preprocessing is stratified by training/test data.

```{r}
dummy_mat <- preprocess_assay_splits(
  lemur_fit,
  "DE",
  transform = "tanh",
  clip_max = "quantile",
  clip_max_value = 0.95,
  clip_noise = "none",
  clip_noise_value = 0.
)
hist(as.vector(dummy_mat$train))
```


# 2. How many factors to fit?

Hard to judge. A few general considerations

- Reasonable reconstruction. Is most of the matrix structure captured?
- Are factors reproducible and exhaustive? Do repeated fit attempts find the same/very similar factors?
- Too many factors should not hurt too much apart from performance. (Look up the theorem prooving that too many factors in NMF assist in finding better minima at lower rank). Should only find finer biclusters, or fit noise-like biclusters that can be removed downstream
- Probably less than LEMUR latent dimension.

The "package" has a small helper function that iterates over proposed factors $k$, and uses a small validation data set to judge reconstruction error over many attempts. It also contains a geometric argument of how similar found factors are -- principal angles between the gene subspaces spanned for different repetitions at fixed $k$. This might aid in guessing a suitable number of factors.

A similar function is available to evaluate the level of sparseness for W and H.

```{r}
dummy_explore <- explore_factors_progress(
  dummy_mat$train,
  c(4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22),
  n_restarts = 10,
  tol = 0.02,
  test_fraction = 0.05
)

print(dummy_explore$plot)

dummy_sparse <- explore_sparsity_progress(
  dummy_mat$train,
  k = 12,
  l1_grid = list(
    c(0, 0),
    c(0.01, 0.0),
    c(0.05, 0.01),
    c(0.10, 0.05),
    c(0.20, 0.10),
    c(0.30, 0.20),
    c(0.30, 0.30),
    c(0.40, 0.40)
  ),
  n_restarts = 10,
  test_fraction = 0.050,
  tol = 0.02
)

print(dummy_sparse$plot)
```


# 3. Fitting NMF

After deciding which factor numbers to try, there is a think wrapper around the `RcppML` package's NMF fitting. It attempts repeated fits and builds a consensus bicluster. Genes and cells are assigned soft memberships to biclusters based on the fraction of repetitions they are loaded into a given factor.

```{r}
support_obj <- build_consensus_nmf(
  X = dummy_mat$train,
  k = 15,
  reps = 20,
  L1 = c(0.1, 0.1),
  zero_tol = 1e-6,
  factor_weight = 0.5,
  seed = 1,
  tol = 0.01
)


```

# 4. Translating soft-membership into hard labels

By defining thresholds for the fraction of recruitments, we turn soft membership into hard labels.

```{r}

res <- get_biclusters(
  support_obj,
  gene_support_threshold = 0.95,
  cell_support_threshold = 0.95
)

res_2 <- apply_sparse_bicluster_thresholds(
  support_obj,
  gene_support_threshold = 0.95,
  cell_support_threshold = 0.8
)
```

# 5. Applying to test data

No longer supported (for now)

```
projections <- .project_biclusters(dummy_fit, dummy_mat$test, recruitment_thrsh_cell)

bic_test <- projections$thresholded
names(bic_test) <- sub("^nmf", "Bic", names(bic_test))
```

We can plot test and training embedding on the full data umap to visually confirm that test data biclusters are compatible with their training counterparts



```{r}
#| fig-width: 12
#| fig-height: 24
#| out-width: 100%

# combine train + test embeddings
all_embedding <- cbind(
  lemur_fit$training_data$embedding,
  lemur_fit$test_data$embedding
)

umap <- uwot::umap(t(all_embedding))

# metadata for all cells
train_df <- as_tibble(lemur_fit$training_data$colData) |>
  mutate(
    dataset = "train",
    cell_id = colnames(lemur_fit$training_data$embedding)
  )

test_df <- as_tibble(lemur_fit$test_data$colData) |>
  mutate(
    dataset = "test",
    cell_id = colnames(lemur_fit$test_data$embedding)
  )

cell_df <- bind_rows(train_df, test_df) |>
  mutate(
    UMAP1 = umap[, 1],
    UMAP2 = umap[, 2]
  )

# plot each bicluster as a facet
plot_df <- bind_rows(lapply(seq_along(res$biclusters), function(i) {
  bic_obj <- res$biclusters[[i]]
  bic_name <- paste0("Bic", i)
  bic_cells <- bic_obj$cell_names

  cell_df |>
    mutate(
      bic = bic_name,
      membership = if_else(cell_id %in% bic_cells, "in bicluster", "other")
    )
}))

ggplot(plot_df, aes(UMAP1, UMAP2)) +

  # background cells first
  geom_point(
    data = subset(plot_df, membership == "other"),
    color = "grey92",
    alpha = 0.5,
    size = 0.2
  ) +

  # bicluster cells on top
  geom_point(
    data = subset(plot_df, membership == "in bicluster"),
    color = "#E64B35",
    alpha = 0.7,
    size = 0.5
  ) +

  facet_wrap(~ bic, ncol = 6) +
  coord_equal() +
  theme_classic()
```



```{r}

pairwise_jaccard <- function(sets) {
  n <- length(sets)

  out <- matrix(
    0,
    nrow = n,
    ncol = n,
    dimnames = list(names(sets), names(sets))
  )

  for (i in seq_len(n)) {
    a <- unique(as.character(sets[[i]]))

    for (j in i:n) {
      b <- unique(as.character(sets[[j]]))

      union_size <- length(union(a, b))

      score <- if (union_size == 0) {
        0
      } else {
        length(intersect(a, b)) / union_size
      }

      out[i, j] <- score
      out[j, i] <- score
    }
  }

  out
}

gene_sets <- lapply(res$biclusters, `[[`, "gene_names")

cell_sets <- lapply(res$biclusters, `[[`, "cell_names")

gene_jaccard <- pairwise_jaccard(gene_sets)

cell_jaccard <- pairwise_jaccard(cell_sets)

pheatmap(gene_jaccard)
pheatmap(cell_jaccard)
```

```{r}
g1 <- res$biclusters[[11]]$gene_names
c1 <- res$biclusters[[11]]$cell_names

sub1 <- assay(lemur_fit, "DE")[g1, c1, drop = FALSE]

irlba::irlba(sub1, nv = 1, nu = 0)$d^2/norm(sub1, type="F")^2

```

# 6. Downstream?

Different options from here on.

- Enrichment of biclusters with cell annotations/other meta data?
- GSEA can reveal programmatic responses to condition contrast, independently of meta data. Find programmatic responses shared between cell types, as well as subdominant programmes that might get lost in cell type bulk analysis.
- Finding programms and subprogramms. E.g. `sertdiff(bicA, bicB)` can reveal how bicA differs from bicB when they share significant overlap -> Programmes and Subprogrammes in DE

